---
title: "F1 World Drivers' Championship"
output: rmarkdown::html_document
author: Camilla Yu
---

### Data

```{r}
library(tidyverse)
library(ggplot2)
library(viridis)
library(ggpubr)
f1 = readRDS(file="data/f1.rds")
```

<br/>

### Task 1 - Tidy the data

#### Write up

After viewing the original data using `View` function, I found that all the 
variables that we are interested in are inside of the `Race` list. So, I 
selected the list `Race` using dollar sign. Now, we need to expand wider and also
longer to find those variables, so I chose to use `hoist` function to rectangle 
the subset of the data and selecting some specific columns. By viewing the 
original data, we could directly extract `race_name`, `round`, `date` variables
since they are on the next level of the list inside of `Race` list and they 
are not list types. And there are also some variables are inside of the 
`Results` list, so at first we need to store the `Results` list into a `result` 
variable, and then use `unnest_longer` function to expand all the 
lists(observations) inside of the `Results` list. After that, we could use 
`hoist` function again to directly extract `position`, `points` and `positionText` 
variables. And because `givenName`, `familyName` and `constructor` variables are 
inside of another `Driver` or `Constructor` list, we need to use `c()` to specify 
the next list we need to look at and then extract the variables we are 
interested in. After extracting all the variables, we need to create a new 
variable `driver` using the value of `givenName` and `familyName` by pasting 
them together. And for `position` variable, we need to check the `positionText` 
first to make sure that the driver actually finished the race and then assign the 
value of `position` to the variable `position` or assign NA otherwise. 
What's more, we need to transfer the data type for `round` and `points` to 
integer type, and transfer `date` to date type. Finally, we select all the 
variables we are interested in and assign it a new name `f1_data` and print out 
the first 10 rows of the data.

#### Code

```{r task1}
f1_data <- tibble::tibble(
  #Select the specific list 'Race' that contains variables we are interested in
  Race = f1$MRData$RaceTable$Races
) %>% 
  #Pipe the target list to hoist function to rectangle the subset of the data
  hoist(
    Race,
    race_name = "raceName",
    round = "round",
    date = "date",
    results = "Results"
  ) %>% 
  #Expand all the lists(observations) inside of the 'results' list 
  select(-Race) %>%
  unnest_longer(results) %>%
  hoist(results,
        position = "position",
        points = "points",
        position_text = "positionText",
        #Extract 'givenName' from list 'Driver'
        givenName = c("Driver", "givenName"),
        familyName = c("Driver", "familyName"),
        constructor = c("Constructor", "name")
  ) %>%
  #Create 'driver' variable by pasting given name and family name together
  mutate(driver = paste(givenName, familyName),
         #Fill the variable 'position' using NA if the driver didn't finish the 
         #race by checking the position_text 
         position = ifelse(position_text == position, 
                           as.integer(position), NA),
         #Transfer 'round' and 'points' variables to integer type 
         round = as.integer(round),
         points = as.integer(points),
         #Transfer 'date' variable to date type
         date = as.Date(date)) %>%
  #Remove uninterested columns from the dataset
  select(-c(givenName, familyName, position_text, results))
#Show the first 10 rows
head(f1_data, n =10)
```

<br/>

### Task 2 - Drivers' Championship

#### Write up

At first, I focused on calculating the total points earned from all rounds for 
each driver. Because we are interested in the total points instead of the points 
for each round, we need to group `driver` variable first to combine all the 
information together for each driver so that we could have distinct names for 
each driver. Then we could use `summarise` function to sum all the points earned
from each round together for each driver as total points. Then, by arranging 
the values for total points in a decreasing order, we could find out who are the 
top drivers. 

Then I focused on the position for each round for each driver. By extracting 
values from position as the value for variables extracted from round, we could
get 17 different columns representing each round and the corresponding position
values for each round. Finally, I combined two data sets together and the order
is arranged by total points in a decreasing order. 

#### Code

```{r task2}
total_points <- f1_data %>%
  #Group by driver to combine information/points together for each driver
  group_by(driver) %>%
  #Sum all the points got from all rounds for each driver 
  summarise(points_total = sum(points)) %>%
  #Arrange the total points in a decreasing order 
  arrange(desc(points_total))
finish_position <- f1_data %>%
  select(driver, round, position) %>%
  #Extract values from position as the value of variables extracted from round
  pivot_wider(., id_cols = driver, 
              names_from = round,
              values_from = position)
#Use left_join function to combine two data sets together 
championship <- left_join(total_points, finish_position, by = "driver")
championship
```

<br/>

### Task 3 - Cumulative Constructors

#### Write up

At first, we need to group by `race_name`, `constructor` and `round` to combine
information together for each constructor in each round. Then, we first sum all
the points earned by each constructor for each round and then arrange the order
for round from the earliest to oldest to make sure we could calculate for the 
cumulative points. Then, we need to group by `constructor` to calculate 
cumulative points earned by each round for each constructor. Finally, I extract 
values from the cumulative points as the value for variables extracted from 
`race_name` as the name for each column, and arrange the value for the last 
column in a decreasing order.

#### Code

```{r task3}
constructor_table <- f1_data %>%
  group_by(race_name, constructor, round) %>%
  #Calculate total points earned for each round for each constructor
  summarise(points2 = sum(points), .groups = 'drop') %>%
  #Arrange the order of round/race_name from earliest to oldest 
  arrange(round) %>%
  group_by(constructor) %>%
  #Calculate the cumulative points earned for each constructor 
  mutate(sum_points = cumsum(points2)) %>%
  #Extract values from cumulative points as the value for variables extracted 
  #from race_name
  pivot_wider(., id_cols = constructor, 
              names_from = race_name,
              values_from = sum_points) %>%
  #Arrange the value from the last column in a decreasing order 
  arrange(desc(.[, dim(.)[2]]))
constructor_table
```

<br/>

### Task 4 - Visualization

#### Write up

To better show the overall performance for each constructor and driver over the 
course of the 2020F1 season, I chose to use cumulative points for each round as 
y-axis and round values as x-axis so that I could identify the top drivers as 
well as top constructors. And for constructors, because there are only 10 of 
them, I use `facet` to show separate plots for each constructor so that I could
compare their overall performance side by side, and also pick the top constructor 
by comparing the value for cumulative points at the final round. For drivers,
since there are 23 of them, it's better to show the subset of them instead of all.
So, I pick the top 5 drivers to visualize their cumulative points earned for each
round to see their overall performance as well as their points trend as the change
of time(round).

From the plot of constructors, Mercedes has the highest cumulative points after 
the round of 10 and has the steepest positive slop, which means Mercedes earned 
more and more points after each round. Red Bull has the second highest cumulative
points after around round 15. The slop is smaller at the first several rounds and
then begins to increase rapidly. The value and trend for cumulative points for 
McLaren and Racing Point are pretty similar, and they have higher points than 
Renault. In general, Mercedes, Red Bull, McLaren, Racing Point and Renault are 
top 5 constructors, and Mercedes really stands out and has weigh higher points
than the rest of them. 

From the plot of drivers, Lewis has the highest cumulative points after round 4.
Valtteri and Max have similar cumulative points for each round, and Valtteri is 
a little bit better than Max. Sergio and Daniel have even similar cumulative points
for each round and Sergio is better than Daniel. In general, Lewis, Valtteri,
Max, Sergio and Daniel are top 5 drivers, and Lewis really stands out and the 
cumulative points are weigh higher than the rest of them, while there is fierce 
competition between Valtteri and Max, as well as between Sergio and Daniel. 

#### Code

```{r task4}
constructor_table2 <- f1_data %>%
  select(constructor, round, points) %>%
  group_by(constructor) %>%
  #Calculate the cumulative points earned for each round for each constructor
  mutate(points2 = cumsum(points)) %>%
  arrange(round)
constructor_plot <- ggplot(constructor_table2) + 
  #Use round as x values and cumulative points for each round as y values 
  geom_point(aes(x = round, 
                 y = points2,
                 #Coloring different constructors 
                 col = constructor)) +
  labs(title = "Cumulative points earned for each round for each constructor",
       x = "Round",
       y = "Points") + 
  scale_color_viridis_d() + 
  #Create separate plots for each constructor 
  facet_grid(.~ constructor) +
  guides(color = "none")
#Find the cutoff value for top 5 total points as the threshold 
threhold <- total_points$points_total[5]
driver_table2 <- f1_data %>%
  select(driver, round, points) %>%
  group_by(driver) %>%
  #Calculative the cumulative points earned for each round for each driver 
  mutate(sum_points = cumsum(points),
         total_points = sum(points)) %>%
  #Filter the top 5 drivers 
  filter(total_points >= threhold)
driver_plot <- ggplot(driver_table2, 
                      #Use round as x values and cumulative points for each 
                      #round as y values 
                      aes(x = round, 
                          y = sum_points,
                          col = driver)) + 
  geom_point() + 
  geom_line() +
  labs(title = "Cumulative points earned for each driver for each round",
       x = "Round",
       y = "Cumsum points")
#Put two plots together 
ggarrange(constructor_plot, driver_plot, ncol = 1, nrow = 2)
```

<br/>
